---
title: "PSY40760 — Week 2 Practical: Getting Started with R"
author: "Your name here"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Before you start

This is an **R Markdown notebook** — a mix of explanatory text (like this) and grey "code chunks" you can run. To run a chunk, click the green ▶ arrow in its top-right corner, or put your cursor inside it and press `Ctrl+Enter` (Windows/Linux) or `Cmd+Enter` (Mac). Try it now:

```{r hello}
print("Hello, R!")
```

You should see the output appear directly beneath the chunk. That's the whole workflow for this module: read, run, edit, run again.

**Where this file lives matters.** Open this project by double-clicking `psy40760-practicals-2026.Rproj` in RStudio — that sets your **working directory** automatically to the repo folder, which is why the file paths in this notebook (like `data/trees.csv` below) will just work. If you open the `.Rmd` file directly instead of opening the `.Rproj` first, paths will break — more on this in the data import section.

Work through this notebook top to bottom. Sections marked **Your turn** are for you to complete — there's no separate answer sheet released before the session, so give them a real go before asking Cameron.

---

# RStudio orientation

When you open RStudio you'll see up to four panes:

- **Source** (top-left) — this file. Where you write and save code.
- **Console** (bottom-left) — where code actually runs, one line at a time. Anything you type here is *not* saved when you close R.
- **Environment/History** (top-right) — every object (variable, dataset, function) you've created this session.
- **Files/Plots/Packages/Help** (bottom-right) — file browser, plot output, installed packages, and documentation.

The critical habit: **write code in the Source pane (this file), not the Console.** The console is for quick one-off checks; the script is your permanent, re-runnable record of what you did — which matters a lot once you're submitting assessed work.

A line starting with `#` is a comment — R ignores it. Use comments constantly to explain *why* you did something to your future self.

```{r comment-demo}
# this is a comment - R skips this line entirely
5 + 5 # you can also comment at the end of a line of code
```

---

# R as a calculator

R evaluates arithmetic exactly the way you'd expect, following standard operator precedence (BODMAS/PEMDAS):

```{r arithmetic}
2 + 6 * 8
(2 + 6) * 8
10 / 3
10 %% 3   # remainder (modulo)
2 ^ 5     # exponent
```

Whitespace almost never matters in R — `2+6` and `2 + 6` give the same answer. Use spaces generously; it makes code easier to read, which matters when someone else (a marker, a collaborator, future-you) has to read it.

The `[1]` in front of your output isn't part of the answer — it's R telling you "this is the first element of the result." You'll understand why once we get to vectors below.

**A common beginner trap:** if you run an incomplete expression, R shows a `+` prompt in the console and waits for you to finish it, rather than erroring immediately.

```r
> 2 +
+
```

If you get stuck in this state, press `Esc` to cancel and start again.

## Your turn

Calculate: what is `2` to the power of `6`, minus the remainder of `17` divided by `5`?

```{r arithmetic-exercise}
# your code here

```

---

# Functions and getting help

A function takes input (arguments) and returns output. `sqrt()` and `log()` are two you'll use constantly:

```{r functions}
sqrt(81)
log(100)        # natural log by default
log(100, base = 10)   # named argument overrides the default
```

**Every function has documentation.** Run `?functionname` or `help(functionname)` to open it in the Help pane. Get comfortable reading these — they follow a consistent structure (Description, Usage, Arguments, Examples) and are the fastest way to answer "wait, how does this argument work?" without leaving RStudio.

```{r help-demo, eval=FALSE}
?log
help(log)
```

## Your turn

Look up `?round`. Use it to round `3.14159` to 2 decimal places (hint: check the `digits` argument).

```{r help-exercise}
# your code here

```

---

# Assignment: giving things names

`<-` assigns a value to a name. Once assigned, that object persists in your Environment for the rest of the session (or until you remove it).

```{r assignment}
my_age <- 25
my_age

reaction_time_ms <- 342.7
reaction_time_ms * 2
```

Naming rules: must start with a letter, can contain letters/numbers/`.`/`_`, and R is **case-sensitive** — `Score` and `score` are different objects. Pick a naming style (e.g. `snake_case`) and stick to it.

```{r rm-demo}
rm(my_age)   # deletes the object from your environment
```

## Your turn

Assign the value `18` to a variable called `n_participants`. Then create a second variable, `n_excluded`, equal to `2`. Calculate and print how many participants remain.

```{r assignment-exercise}
# your code here

```

---

# Data types

Every value in R has a **class** — the three you'll meet constantly are:

```{r types}
class(4.5)        # "numeric"
class("hello")     # "character" - note the quotes
class(TRUE)        # "logical" - TRUE/FALSE, always capitalised, never quoted
```

A classic early mistake: typing `true` instead of `TRUE`, or `"TRUE"` (a character string) instead of `TRUE` (a logical value) — these are three different things to R.

---

# Vectors

A vector is an ordered collection of values of the *same* type — the fundamental building block of almost everything else in R.

```{r vector-creation}
1:10                        # a sequence
c(3, 7, 2, 9)                # combine arbitrary values
rep(1:3, times = 2)          # repeat the whole sequence
rep(1:3, each = 2)           # repeat each element
```

Functions apply across a whole vector at once — no loop needed:

```{r vector-math}
scores <- c(14, 18, 9, 22, 16)
mean(scores)
sd(scores)
scores * 2          # every element doubled
```

**Subsetting** — pulling specific elements out with `[ ]`:

```{r subsetting}
scores[1]           # first element
scores[c(1, 3)]      # first and third
scores[-1]           # everything except the first
scores[scores > 15]  # logical subsetting - only values above 15
```

## Your turn

Using `rep()`, reproduce the sequence `2 2 2 5 5 5 8 8 8` in one line. Then use `rep()` again to reproduce `2 5 8 2 5 8 2 5 8`. Think about which argument controls which pattern before you try.

```{r vector-exercise}
# your code here

```

---

# Logical operators

```{r logicals}
5 > 3
5 == 5    # equality - two equals signs, NOT one
5 != 4    # not equal
```

Combining conditions — think of `&` and `|` as needing a **complete, standalone statement on each side**, not a shared subject:

> Imagine asking "is it Bob AND Sally who are coming to the party?" You have to name *both* people fully — `bob_coming & sally_coming` — you can't just say "Bob and Sally" and have R figure out you meant two separate checks.

```{r logical-combine}
age <- 22
has_consent <- TRUE

age >= 18 & has_consent     # both must be TRUE
age < 18 | !has_consent     # either can be TRUE
```

A common mistake that trips this up:

```{r logical-trap, eval=FALSE}
pet <- "cat"
pet == "cat" | "dog"    # WRONG - does not check pet == "dog"
pet == "cat" | pet == "dog"  # correct - full statement each side
```

## Your turn

Without running it first, predict what each of these will return, then check yourself:

```{r logical-exercise}
!TRUE
(TRUE & FALSE) | TRUE
(5 > 3) & (2 == 3)
```

---

# Data frames

A data frame is a table: rows are observations, columns are variables. R ships with several built-in ones for practice — we'll use `trees` (diameter, height and volume of 31 black cherry trees).

```{r dataframe-explore}
head(trees)          # first 6 rows
str(trees)            # structure: types of each column
summary(trees)         # quick descriptive stats
```

Subsetting a data frame uses `df[rows, columns]` — leave one side blank to mean "all":

```{r dataframe-subset}
trees[1, ]             # first row, all columns
trees[, "Height"]       # Height column, all rows
trees$Height             # same thing, using $
trees[trees$Height > 80, ]   # only rows where Height exceeds 80
```

## Your turn

How many trees in the dataset have a `Volume` greater than 40? (Hint: combine logical subsetting with a function that counts things — `sum()` on a logical vector counts the `TRUE`s.)

```{r dataframe-exercise}
# your code here

```

---

# Importing your own data {#importing-data}

Everything so far has used data already living inside R. In practice you'll almost always start from a **file someone gives you** — a `.csv` from a lab, a survey export, a public dataset. This section is the one most beginners get stuck on, so we're doing it properly, step by step.

## Working directory

R always has a "working directory" — the folder it looks in by default when you ask it to open a file. If you opened this project via `psy40760-practicals-2026.Rproj`, your working directory is already set correctly. Confirm it:

```{r getwd}
getwd()
```

This should print a path ending in `.../psy40760-practicals-2026`. If it doesn't, close this file, then re-open the project by double-clicking the `.Rproj` file — don't just open the `.Rmd` on its own.

**Why this matters:** a command like `read.csv("data/trees.csv")` is a *relative* path — R looks for a `data` folder *inside whatever your working directory currently is*. If your working directory is wrong, the exact same command fails on your machine even though it works on someone else's. This is the single most common "but it worked for my classmate!" bug in this module.

## Reading a CSV

```{r read-csv}
trees_imported <- read.csv("data/trees.csv")
head(trees_imported)
str(trees_imported)
```

Compare `trees_imported` to the built-in `trees` you explored earlier in `data/trees.csv` — same data, but now you've taken it through the exact journey a real dataset takes: a file on disk, into an R object, via a working directory that has to be correct.

## When it goes wrong

You will, at some point, see this error:

```
Error in file(file, "rt") : cannot open the connection
```

This almost always means one of:

1. **Wrong working directory** — run `getwd()` and check it matches where the file actually is (use the Files pane, bottom-right, to look).
2. **Typo in the filename or path** — R is case-sensitive and exact; `Data/Trees.csv` is not the same as `data/trees.csv` on most systems.
3. **File extension mismatch** — check the file really is `.csv` and not `.csv.txt` or similar (file managers sometimes hide the real extension).

There is no shortcut around this — every R user hits this error regularly. The fix is always: check `getwd()`, check the Files pane, check spelling.

## Your turn

The `data/` folder also contains nothing else right now, but imagine a classmate emails you a file called `Survey_Data.CSV` and says "just read it in!" Write the line of code you'd use, and in a comment, note the two most likely reasons it might fail on your machine even if it worked on theirs.

```{r import-exercise}
# your code here, plus a comment noting two likely failure reasons

```

---

# Lists (a quick look)

A list can hold *different* types of things together — including vectors of different lengths, or even other data frames.

```{r lists}
participant <- list(id = 1, initials = "AB", scores = c(14, 18, 22))
participant$scores       # access by name
participant[["initials"]] # equivalent, double-bracket form
```

You'll meet lists properly once we start fitting statistical models later in the module — a fitted model in R is, under the hood, a list.

---

# Matrices (a quick look)

A matrix is like a data frame but every value must be the *same* type (usually all numeric).

```{r matrices}
m <- matrix(1:6, nrow = 2)
m
colnames(m) <- c("A", "B", "C")
m
```

---

# Wrap-up

You've covered: the RStudio interface, arithmetic and functions, assignment, the three basic data types, vectors and subsetting, logical operators, data frames, importing a real CSV file (properly, including what to do when it breaks), and a first look at lists and matrices.

**Before next week:** if any "Your turn" section gave you trouble, that's exactly the kind of thing to bring to Cameron or Ondrej — better now than in Week 4 when we start layering the tidyverse on top of all of this.
